A Wait-free Queue with Polylogarithmic Step Complexity

### 2022-12-31 podc_paper/related.tex
formatting

\paragraph{List-based Queues}
The MS-queue \cite{MS98} is a lock-free queue that has stood the test of time.
A version of it is
included in the standard Java Concurrency Package.  %java.util.concurrent.ConcurrentLinkedQueue
See the paper that introduced it for a survey of the early history of concurrent queues.
As mentioned above, the MS-queue suffers from the CAS retry problem because of contention at the front and back of the queue.
Thus, it is lock-free but not wait-free and has an amortized\Hossein{Maybe we can say worst-case. I see that you have used amortized step complexity consistently everywhere but maybe it is fair to say that these papers were not designed to improve the worst-case amortized step complexity.}
\Eric{Amortized step complexity is defined in first parag of paper to be worst case over all executions of the average number of steps per operation, so I think the worst-case aspect of it is clear.  Also, if we just said ``worst-case'' without ``amortized'' it would be confused with worst-case of an individual operation, which is infinite in this case.  By the way, saying that previous implementations have $\Omega(p)$ step complexity, even in the amortized sense, makes the new contribution even more of a contrast.}
step complexity of $\Theta(p)$ per operation.

@@

\paragraph{Array-Based Queues}
Arrays can be used to implement queues with bounded capacity \cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01}.  
%Similar to the front and back pointers of list-based queues, indices
Dequeues and enqueues update
indexes of the front and back elements using CAS instructions.
Gidenstam, Sundell, and Tsigas~\cite{DBLP:conf/opodis/GidenstamST10} avoid
the capacity constraint by using a linked list of arrays.
These solutions also use $\Omega(p)$ steps per operation due to the CAS retry problem.

@@

\paragraph{Universal Constructions}
One can also build a queue using a universal construction \cite{10.1145/114005.102808}.
Jayanti \cite{DBLP:conf/podc/Jayanti98a} observed that
the universal construction of Afek, Dauber, and
Touitou~\cite{DBLP:conf/stoc/AfekDT95} can be modified to use $O(\log p)$ steps per operation, 
assuming that words can store $\Omega(p \log p)$ bits. 
(Thus, in terms of operations on reasonably-sized $O(\log p)$-bit words, their construction would take $\Omega(p\log p)$ steps per operation.)
Fatourou and Kallimanis \cite{FK14} used their universal construction based on fetch\&add and LL/SC instructions
to implement a queue, but its step complexity is also $\Omega(p)$.
\here{Double check this.}

\paragraph{Restricted Queues}
David introduced the first sublinear-time queue
\cite{DBLP:conf/wdag/David04}, but it works only for a single enqueuer.
It uses fetch\&increment and swap  instructions and takes $O(1)$ steps per operation, but
uses unbounded memory.  Bounding the space increases the steps per operation to $\Omega(p)$.

@@

\paragraph{Other Primitives}
Khanchandani and Wattenhofer \cite{KW18} gave a wait-free queue implementation
with $O(\sqrt{p})$ step complexity using non-standard synchronization primitives
called half-increment and half-max, which can be viewed as particular kinds of
double-word read-modify-write operations.
They use this as evidence that their primitives can be more efficient than CAS
since previous CAS-based queues all required $\Omega(p)$ step complexity.
Our new implementation counters this argument.

\paragraph{Fetch\&Increment Objects}
There are sublinear-time implementations of another object
with consensus number 2.
Two papers \cite{ERW12,10.1007/978-3-642-41527-2_20}
implemented fetch\&increment objects from LL/SC objects
using a polylogarithmic number of steps per operation.
Our approach is similar to that of Ellen, Ramachandran and Woelfel \cite{ERW12},
which uses a tournament tree to keep track of the fetch\&increments
that have been performed, as in the universal construction of \cite{DBLP:conf/stoc/AfekDT95}.
However, our construction requires more intricate
data structures to represent sets of operations, since the state of a queue cannot easily be represented as succinctly
as the single-word state of a fetch\&increment object.

>>>>

\emph{List-based Queues.}
The MS-queue \cite{MS98} is a lock-free queue that has stood the test of time.
A version of it is
included in the standard Java Concurrency Package.  %java.util.concurrent.ConcurrentLinkedQueue
See the paper that introduced it for a survey of the early history of concurrent queues.
As mentioned above, the MS-queue suffers from the CAS retry problem because of contention at the front and back of the queue.
Thus, it is lock-free but not wait-free and has an amortized\Hossein{Maybe we can say worst-case. I see that you have used amortized step complexity consistently everywhere but maybe it is fair to say that these papers were not designed to improve the worst-case amortized step complexity.}
\Eric{Amortized step complexity is defined in first parag of paper to be worst case over all executions of the average number of steps per operation, so I think the worst-case aspect of it is clear.  Also, if we just said ``worst-case'' without ``amortized'' it would be confused with worst-case of an individual operation, which is infinite in this case.  By the way, saying that previous implementations have $\Omega(p)$ step complexity, even in the amortized sense, makes the new contribution even more of a contrast.}
step complexity of $\Theta(p)$ per operation.

@@

\emph{Array-Based Queues.}
Arrays can be used to implement queues with bounded capacity \cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01}.  
%Similar to the front and back pointers of list-based queues, indices
Dequeues and enqueues update
indexes of the front and back elements using CAS instructions.
Gidenstam, Sundell, and Tsigas~\cite{DBLP:conf/opodis/GidenstamST10} avoid
the capacity constraint by using a linked list of arrays.
These solutions also use $\Omega(p)$ steps per operation due to the CAS retry problem.

@@

\emph{Universal Constructions.}
One can also build a queue using a universal construction \cite{10.1145/114005.102808}.
Jayanti \cite{DBLP:conf/podc/Jayanti98a} observed that
the universal construction of Afek, Dauber, and
Touitou~\cite{DBLP:conf/stoc/AfekDT95} can be modified to use $O(\log p)$ steps per operation, 
assuming that words can store $\Omega(p \log p)$ bits. 
(Thus, in terms of operations on reasonably-sized $O(\log p)$-bit words, their construction would take $\Omega(p\log p)$ steps per operation.)
Fatourou and Kallimanis \cite{FK14} used their universal construction based on fetch\&add and LL/SC instructions
to implement a queue, but its step complexity is also $\Omega(p)$.
\here{Double check this.}

\emph{Restricted Queues.}
David introduced the first sublinear-time queue
\cite{DBLP:conf/wdag/David04}, but it works only for a single enqueuer.
It uses fetch\&increment and swap  instructions and takes $O(1)$ steps per operation, but
uses unbounded memory.  Bounding the space increases the steps per operation to $\Omega(p)$.

@@

\emph{Other Primitives.}
Khanchandani and Wattenhofer \cite{KW18} gave a wait-free queue implementation
with $O(\sqrt{p})$ step complexity using non-standard synchronization primitives
called half-increment and half-max, which can be viewed as particular kinds of
double-word read-modify-write operations.
They use this as evidence that their primitives can be more efficient than CAS
since previous CAS-based queues all required $\Omega(p)$ step complexity.
Our new implementation counters this argument.

\emph{Fetch\&Increment Objects.}
There are sublinear-time implementations of another object
with consensus number 2.
Two papers \cite{ERW12,10.1007/978-3-642-41527-2_20}
implemented fetch\&increment objects from LL/SC objects
using a polylogarithmic number of steps per operation.
Our approach is similar to that of Ellen, Ramachandran and Woelfel \cite{ERW12},
which uses a tournament tree to keep track of the fetch\&increments
that have been performed, as in the universal construction of \cite{DBLP:conf/stoc/AfekDT95}.
However, our construction requires more intricate
data structures to represent sets of operations, since the state of a queue cannot easily be represented as succinctly
as the single-word state of a fetch\&increment object.
